\section{Chapter 1: Overview of Some Basic Concepts}

\subsection{Operator v Statement}

An operator is a symbol that acts of data in some way. For example +, -, \% etc.

A statement is a complete instruction that tells a program to do something e.g.
an if statement, a while statement etc.

\subsection{Strongly Typed v Weakly Typed Language}
A strongly typed language is one whereby the type of every value is strictly enforced, and does
not allow one to mix types without an explicit conversion. Ada is a strongly typed language. C is not.

\begin{lstlisting}[caption={Example of Strongly Typed Language}, label={code:stl}]
#include <stdio.h>

x = float
y = string

z = x + y \\ Error in STL

z = x + float(y) \\Acceptable in most STL
\end{lstlisting}

\subsection{Static v Dynamic Typed}

\textbf{Static} means that when a variable is assigned a type, it keeps that type and its type is checked at
compile time. Example languages are C, Rust, Go and Ada

\begin{lstlisting}[caption={Example of Static Typed Language}, label={code:static}]
int x = 5;
x = "hello";   // compile-time error
\end{lstlisting}

\noindent So, strongly typed --> cannot mix variable types and operations

\noindent Statically typed --> Once a variable is assigned a type it maintains that type and 
this is checked at compile time.

\textbf{Dynamically typed} means a varibale can be assigned one type somewhere, and re-assigned 
another type later on. Example languages are Python, JavaScript, Ruby.

\begin{lstlisting}[caption={Example of Static Typed Language}, label={code:static}]
int x = 5;
x = "hello";   // NO compile-time error
\end{lstlisting}

\subsection{Procedural v Functional Language}

\textbf{Procedural} languages are based around writing functions that operate on data. Examples are C,
Python, Ada. Use looping structures, and can be viewed as carrying out programs step-by-step.

\noindent \textbf{Functional} languages are based on mathematical functions and expressions. Do not have, 
e.g., for loops, instead using recursion. Example languages are Haskell, OCaml. They are less step-by-step
and more based around writing and calling functions which are the mian building blocks.

\subsection{Imperative v Declarative Programming}

\textbf{Imperative} programming focuses on how to do something by providing step-by-step instructions.
Example languages are C, Python, Ada.

\noindent \textbf{Declarative} programming focuses on what the desired outcome is without specifying 
the exact steps e.g. Coq (Gallina), SQL, Haskell.

\noindent Difference between i-- and --i:

\begin{lstlisting}[caption={Pre- vs Post-Decrement}, label={code:dec_ops}]
int i = 5;

int a = i--;   // a = 5, i becomes 4 (post-decrement)
int b = --i;   // i becomes 3, b = 3 (pre-decrement)
\end{lstlisting}


